Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-29 setup.tex
edits to section 4 with corresponding tweaks to appendix A.  General editing for clarity and conciseness.  Also attempted to blunt perceived criticism of VGSB (which contains many aspects) by focusing on the 'fully radially local approximation'.
more edits to 3.3 will require merge

\subsubsection{Density structure}
The equilibrium density field follows from vertical hydrostatic equilibrium (Eq.\ \ref{vert_equilibrium}).
The solution for $\Gamma\neq1$ is
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right).
\end{align}
Note that for $\Gamma > 1 +  h^2$ there is a disk surface $H_s$
where $\rho(r,H_s)=0$. 
% Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
% $\rho(r,H)=0$. If we parametrize
% \begin{align}
%   H(r)\equiv h r,
% \end{align}
% then the imposed sound-speed and disk thickness are related by 
% \begin{align}
%   h^2(r) = \left[1-\frac{ h^2(r)}{\Gamma-1}\right]^{-2}-1. 
% \end{align}

The vertically isothermal case, $\Gamma = 1$, can be found either 
independently or as a limit of Eq. \ref{eqm_dens} to be 
\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},   \\
  & \approx \exp {\left[ - {z^2 \over 2 H^2}\right]} \label{rhoisothin},
  \end{align}
  where Eq.\ \ref{rhoisothin} takes the $z \ll r$ thin disk limit.

 \subsubsection{Rotation and vertical shear profiles}
The equilbrium rotation field follows from the density field and centrifugal balance (Eq.\ \ref{hori_equilibrium}),
giving 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2(r,z)}{\OmK^2(r)}=1 +
  \left(\frac{p+q}{\Gamma}\right) h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
for all $\gamma$, where $s\equiv q+p(1-\Gamma)$.

Notice that for $\Gamma = 1$, $s/\Gamma = q$...

...
%\subsubsection{Vertically isothermal disks}
%For $\Gamma=1$ the equilibrium density and rotation profiles are
%given, respectivley, by 
%\begin{align}
%  \frac{\rho(r,z)}{\rho_0(r)} &=
%  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},\\    
%  \frac{\Omega^2(r,z)}{\OmK^2(r)}& =1+ (p+q) h^2(r) + q\left(1 -
%    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
%\end{align}
%which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
%taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
%vertically isothermal disks which is the relevant case for passively irradiated
%PPDs \citep{chiang97}.

\subsection{Vertical shear}\label{vshear_def}
Note that the disk is in general baroclinic with $\Omega =
\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot} and Eq. \ref{vertiso_eqm}, but is also 
implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
Eq. \ref{eqm_press},  
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
\end{align}
For the above disk models,
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz\OmK^2}{\Gamma
    r\left(1+z^2/r^2\right)^{3/2}}. \label{vertical_shear_ex} 
\end{align}
Note that at fixed $r$ there is a maximum vertical shear rate at $|z|=r/\sqrt{2}$,
which is expected to limit the growth rate of instabilities associated with vertical shear.

\subsection{Entropy gradients and vertical buoyancy}
The specific entropy is defined as $S\equiv
c_p\ln{P^{1/\gamma}/\rho}$, where $c_p$ is the heat capacity at
constant pressure. For our disk models,
\begin{align}
  &\frac{\p S}{\p r} = c_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = c_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z},\label{dSdz}
\end{align} 
which are used below.

For discussion purposes it is useful to define the vertical buoyancy
frequency $N_z$ via 
\begin{align}
  N_z^2 \equiv - c_p^{-1}g_z \frac{\p S}{\p z} = c_s^2 \left(\frac{\gamma -
      \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,   \label{nzsq_def}
\end{align}
where $g_z\equiv \rho^{-1}\p_zP $. The second equality applies to the
disk equilibria described above (using Eq. \ref{dSdz}). We are
interested in sub-adiabatically stratified disks with $\Gamma <
\gamma$ so that $N_z^2\geq0$. % , and the disk is 
% convectively stable in the vertical direction.

\subsection{Stability against axisymmetric adiabatic
  perturbations}\label{solberg}
A rotating flow is stable against axisymmetric adiabatic perturbations
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho c_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum. For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\OmK^2$ and
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O( h^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O( h^2)$. Since we consider
$ h\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
cooling, as shown below.

>>>>

\subsubsection{Density structure}
The equilibrium density field follows from vertical hydrostatic equilibrium (Eq.\ \ref{vert_equilibrium}).
The solution for $\Gamma\neq1$ is
\begin{align}\label{eqm_dens}
  &\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} = 1 +
  \frac{\left(\Gamma-1\right)}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right).
\end{align}
Note that for $\Gamma > 1 +  h^2$ there is a disk surface $H_s$
where $\rho(r,H_s)=0$.  For $h \ll 1$, this surface is at $H_s/H \approx \sqrt{2/(\Gamma-1)}$.
% Note that for $\Gamma\neq1$ the disk has a surface $z=H(r)$ such that
% $\rho(r,H)=0$. If we parametrize
% \begin{align}
%   H(r)\equiv h r,
% \end{align}
% then the imposed sound-speed and disk thickness are related by 
% \begin{align}
%   h^2(r) = \left[1-\frac{ h^2(r)}{\Gamma-1}\right]^{-2}-1. 
% \end{align}

The vertically isothermal case, $\Gamma = 1$, can be calculated either 
as a special case or as the $\Gamma \rightarrow 1$ limit of Eq. \ref{eqm_dens} to be 
\begin{subequations}\begin{align}
  \frac{\rho(r,z)}{\rho_0(r)} &=
  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},   \\
  & \approx \exp {\left[ - {z^2 \over 2 H^2}\right]} \label{rhoisothin},
  \end{align}\end{subequations}
  where Eq.\ \ref{rhoisothin} takes the $z \ll r$ thin disk limit.  We will primarily focus on
vertically isothermal disks, the relevant case for passively irradiated
PPDs \citep{chiang97}.

 \subsubsection{Rotation and vertical shear profiles}\label{vshear_def}
The equilibrium rotation field, $\Omega(r,z)$ follows from the density field and centrifugal balance (Eq.\ \ref{hori_equilibrium}),
giving 
\begin{align}\label{eqm_rot}
  \frac{\Omega^2}{\OmK^2}=1 +
  \frac{p+q}{\Gamma} h^2(r) 
  +\frac{s}{\Gamma} \left(1-\frac{1}{\sqrt{1+z^2/r^2}}\right), 
\end{align}
for all $\Gamma$, where $s\equiv q+p(1-\Gamma)$.

The vertical shear shear follows as
\begin{align}
  r\frac{\p\Omega^2}{\p z} = \frac{sz}{\Gamma r}  \frac{\OmK^2}{
    \left(1+z^2/r^2\right)^{3/2}} . \label{vertical_shear_ex} 
\end{align}
For $\Gamma = 1$, notice that $s/\Gamma = q$ is the only disk parameter that affects vertical shear.
Vertical shear increases linearly with height near the mid plane, and $\p_z \Omega^2$ is maximized at
the smaller of $r/\sqrt{2}$ or $H_s$.

A more general expression for vertical shear holds for vertical polytopes 
(which satisfy Eq.\ \ref{eqm_press}), with no assumed radial structure:
\begin{align}
  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}, \label{vertical_shear}
\end{align}
where $g_z = \p_z P/\rho$.

%\subsubsection{Vertically isothermal disks}
%For $\Gamma=1$ the equilibrium density and rotation profiles are
%given, respectivley, by 
%\begin{align}
%  \frac{\rho(r,z)}{\rho_0(r)} &=
%  \exp{\left[\frac{1}{ h^2}\left(\frac{1}{\sqrt{1+z^2/r^2}}-1\right)\right]},\\    
%  \frac{\Omega^2(r,z)}{\OmK^2(r)}& =1+ (p+q) h^2(r) + q\left(1 -
%    \frac{1}{\sqrt{1+z^2/r^2}}\right),\label{vertiso_eqm}
%\end{align}
%which may be obtained from Eq. \ref{eqm_dens} and Eq. \ref{eqm_rot} by
%taking the limit $\Gamma\to 1$. We will, in fact, primarily focus on
%vertically isothermal disks which is the relevant case for passively irradiated
%PPDs \citep{chiang97}.

%\subsection{Vertical shear}\label{vshear_def}
%Note that the disk is in general baroclinic with $\Omega =
%\Omega(r,z)$. This is explicit from Eq. \ref{eqm_rot} and Eq. \ref{vertiso_eqm}, but is also 
%implied by Eq. \ref{vert_equilibrium}---\ref{hori_equilibrium} and 
%Eq. \ref{eqm_press},  
%\begin{align}
%  r\frac{\p\Omega^2}{\p z} = - \rho^{\Gamma-1}\frac{\p\ln\rho}{\p
%    z}\frac{dK}{dr} = - \frac{g_z}{\Gamma}\frac{d\ln K}{dr}.\label{vertical_shear}
%\end{align}
%For the above disk models,
%\begin{align}
%  r\frac{\p\Omega^2}{\p z} = \frac{sz\OmK^2}{\Gamma
%    r\left(1+z^2/r^2\right)^{3/2}}. \label{vertical_shear_ex} 
%\end{align}
%Note that at fixed $r$ there is a maximum vertical shear rate at $|z|=r/\sqrt{2}$,
%which is expected to limit the growth rate of instabilities associated with vertical shear.

\subsection{Entropy gradients and vertical buoyancy}
The gradients of entropy, $S\equiv
C_p\ln{P^{1/\gamma}/\rho}$, in our disk models are
\begin{subequations}
\begin{align}
  &\frac{\p S}{\p r} = C_p\left[\frac{s}{\gamma r} +
    \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p
       r}\right],\\
  &\frac{\p S}{\p z} = C_p
  \left(\frac{\Gamma}{\gamma}-1\right)\frac{\p \ln{\rho}}{\p z} = {C_p g_z \over c_s^2} {\Gamma - \gamma \over \gamma},\label{dSdz}
\end{align} 
\end{subequations}
 where $C_p$ is the heat capacity at
constant pressure.

The vertical buoyancy frequency $N_z$ is 
\begin{align}
  N_z^2 \equiv - C_p^{-1}g_z \frac{\p S}{\p z} = {g_z^2 \over c_s^2}{\gamma - \Gamma \over \gamma}. %c_s^2 \left(\frac{\gamma -
   %   \Gamma}{\gamma}\right)\left(\frac{\p\ln\rho}{\p z}\right)^2,  
    \label{nzsq_def}
\end{align}
We only consider convectively stable disks with with $\Gamma <
\gamma$ and  $N_z^2\geq0$. % , and the disk is 
% convectively stable in the vertical direction.

\subsection{Stability without cooling}\label{solberg}
In the absence of cooling, axisymmetric stability is ensured
if both Solberg-Hoiland criteria  are satisfied:
\begin{align}
  \kappa^2 - \frac{1}{\rho C_p}\nabla P \cdot \nabla S &> 0,\\
  -\frac{\p P}{\p z}\left(\frac{\p j^2}{\p r}\frac{\p S}{ \p z} -
    \frac{\p j^2}{\p z}\frac{\p S}{\p r} \right)&>0. \label{second_SH} 
\end{align}
\citep{tassoul78}, where $j=r^2\Omega$ is the specific angular
momentum and $\kappa = r^{-3} \p_r j^2 $ is the epicyclic frequency.  For rotationally-supported, thin disks %  pressure gradients are small 
% compared to rotation, so 
the first criterion is generally satisfied. 
Thus, to assess stability, we consider the second
criterion. 
Without loss of generality, we evaluate Eq. \ref{second_SH} in $z>0$
where $\p_zP<0$. Here we approximate $\p_r j^2\simeq r^3\OmK^2$ and
use Eq. \ref{vertical_shear} for $\p_zj^2$. Then 
\begin{align}\label{solberg2}
  % \gamma - \Gamma > \frac{s h^2}{\Gamma}
  % \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s
  %   -\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]. 
  \gamma - \Gamma > \frac{ h^2}{\Gamma}
  \left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left[s^2
    -s\left(\gamma-\Gamma\right)\frac{\p\ln{\rho}}{\p\ln{r}} \right]
\end{align} 
implies stability. For typical model parameters the right hand
side (RHS) is $O( h^2)$. Hence, instability in the adiabatic limit  
requires $\gamma < \Gamma + O( h^2)$. Since we consider
$ h\ll1$, this implies the equilibrium disk must have a small
entropy gradient in order to be unstable to axisymmetric adiabatic
perturbations. This is generally not true for passively irradiated, stably
stratified disks \citep{chiang97}, in which the VSI requires rapid
cooling, as shown below.
